---
title: "Modeling Shot Efficiency in the NBA "
author: "Stat guys: Lewis Eatherton, Team member 2, Team member 3, Team member 4"
date: "10/28/20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)

library(tidyverse)
library(knitr)
library(broom)
library(leaps)
shots <- read_csv("data/shot_logs.csv")
```

Your written report goes here! Before you submit, make sure your code chunks are turned off with `echo = FALSE` and there are no warnings or messages with `warning = FALSE` and `message = FALSE` 

### Introduction and EDA


talk about why we chose this topic and what our expectations our... similar to proposal



```{r}
### Data cleaning
shots <- shots %>%
  mutate(Make = case_when(
    SHOT_RESULT == "made" ~ 1,
    SHOT_RESULT == "missed" ~ 0))

shots <- shots %>%
  mutate(Make = as.factor(Make))
```


```{r}
##Checking conditions

### Linearity
library(Stat2Data)

emplogitplot1(Make ~ SHOT_CLOCK, data = shots, 
              ngroups = 5)

emplogitplot1(Make ~ CLOSE_DEF_DIST, data = shots, 
              ngroups = 10)

emplogitplot1(Make ~ SHOT_DIST, data = shots, 
              ngroups = 10)

emplogitplot1(Make ~ TOUCH_TIME, data = shots, 
              ngroups = 10)


shots %>%
  count(CLOSEST_DEFENDER, Make) %>%
  group_by(CLOSEST_DEFENDER) %>%
  mutate(prop = n/sum(n)) %>%
  filter(Make == "1") %>%
  mutate(emp_logit = log(prop/(1-prop)))


shots_players <- shots %>%
  filter(CLOSEST_DEFENDER == "Leonard, Kawhi"| CLOSEST_DEFENDER ==  "Curry, Stephen")

shots_players <- shots_players %>%
  filter(CLOSE_DEF_DIST < 5)

shots_players <- shots_players %>%
  mutate(Make = as.factor(Make))

shots_players <- shots_players %>%
  mutate(CLOSEST_DEFENDER = as.factor(CLOSEST_DEFENDER))

emplogitplot2(Make ~ CLOSE_DEF_DIST + CLOSEST_DEFENDER, data = shots_players)

```

Linearity is not totally satisfied... graphs aren't linear

independence and randomness seems fine accorind to how data was collected


we should also look into collinearity

### Creating Model
```{r}

full_mod <- glm(Make ~ SHOT_CLOCK + DRIBBLES + TOUCH_TIME + SHOT_DIST + CLOSE_DEF_DIST, data = shots, family = "binomial")

drop1(full_mod)



full_mod <- glm(Make ~ SHOT_CLOCK + DRIBBLES + TOUCH_TIME + SHOT_DIST + CLOSE_DEF_DIST + CLOSEST_DEFENDER, data = shots_players, family = "binomial")

drop1(full_mod)

current <- glm(Make ~ SHOT_CLOCK + TOUCH_TIME + SHOT_DIST + CLOSE_DEF_DIST + CLOSEST_DEFENDER, data = shots_players, family = "binomial")
drop1(current)

current <- glm(Make ~ SHOT_CLOCK + TOUCH_TIME + SHOT_DIST + CLOSE_DEF_DIST, data = shots_players, family = "binomial")
drop1(current)

current <- glm(Make ~ SHOT_CLOCK + SHOT_DIST + CLOSE_DEF_DIST, data = shots_players, family = "binomial")
drop1(current)


### testing for this interaction term

current_interaction <- glm(Make ~ SHOT_CLOCK + SHOT_DIST + CLOSE_DEF_DIST + CLOSE_DEF_DIST * CLOSEST_DEFENDER, data = shots_players, family = "binomial")

anova(current, current_interaction, test = "Chisq") %>%
  tidy()

current%>%
  tidy() %>%
  kable(digits = 3)


```

talk about final model outcome and how we came to it


###Discussion

Talk about our results, the limitations of these results, and what'd we do differently...
